---
title: "Exercise 12"
output:
  pdf_document: default
header-includes:
  - \renewcommand{\P}{\operatorname{P}}
---

```{r, message=FALSE, include=FALSE}
require(UsingR)
```

## 1. Born Babies (T)

On average, the number of babies born in Cleveland, Ohio, in the month of September
is 1472. On January 26, 1977 the city was immobilized by a blizzard.
Nine months later, in September 1977, the recorded number of births was 1718.
Can the increase of 246 be attributed to chance? To investigate this,
the number of births in the month of September is modelled by a Poisson
distributed random variable $T$ with parameter $\mu$, and we test $H_0:\mu=1472$.

$$
\begin{aligned}
  \mu_0 &= 1472 \\
  t &= 1718 \\
  H_0 &: \mu = \mu_0 \\
  H_1 &: \mu > \mu_0
\end{aligned}
$$

(a) What would you choose as the alternative hypothesis? 

$H_1:\mu > 1472$

(b) In which direction do values of $T$ provide evidence against $H_0$ (and in favour of $H_1$)?

Right. Higher numbers for the mean provides evidence against $H_0$ and in favour of $H_1$  
Number to the left also provide evidence against $H_0$ but even more against $H_1$.

(c) Compute the $p$-value corresponding to $t=1718$. 

<!-- $$
\begin{align*}
  p &= \P(\mu \geq 1718) \\
  &= `r 1 - ppois(1718, lambda=1472)`
\end{align*}
$$ -->

$$
\begin{aligned}
  p &= \P(\mu \geq 1718) \\
  &= `r pnorm(1718, mean = 1472, sd = sqrt(1472), lower.tail = FALSE)`
\end{aligned}
$$

Calculating the upper-tail p-value in R:  
`#!r pnorm(1718, mean = 1472, sd = sqrt(1472), lower.tail = FALSE)`

(d) What is your conclusion about whether the increase in births was attributed to chance?

The p-value is very low so there is grounds for rejecting the null hypothesis
in favor of the alternative hypothesis.

Hint: you may use the fact that the Poisson distribution of $T$ can be approximated by an $N(\mu,\mu)$ distribution. 

## 2. Type I and II errors (T)

Let $T$ be a random variable following an $N(\mu,1)$ distribution. Assume testing the hypothesis $H_0: \mu=0$ against the alternative hypothesis $H_1: \mu \neq 0$ using the test statistic $T$.

(a) With the decision to reject the null hypothesis if the realization $|t|>1$ what is the probability of committing a type I error?

Find two-tailed p-value for $|t| > 1$

$$
\begin{aligned}
  p &= \P(|t| > 1) = 2\P(t < -1) \\
  &= `r 2 * pnorm(-1, mean=0, sd=1)`
\end{aligned}
$$

(`#!r 2 * pnorm(-1, mean = 0, sd = 1)`)

(b) Assuming that the true value of $\mu$ is 1, what is the probability of committing a type II error? 

$$
\begin{aligned}
  \P(|t| < 1) &= \P(-1 < t < 1) = \P(t < 1) - \P(t < -1) \\
  &= `r pnorm(1, mean=1, sd=1) - pnorm(-1, mean=1, sd=1)`
\end{aligned}
$$

(`#!r pnorm(1, mean=1, sd=1) - pnorm(-1, mean = 1, sd = 1)`)


## 3. Coin flipping (T)

Assume that you flipped a coin $N=11$ times. You got tails 7 times and heads 4 times, after which the coin fell into a well and you lost it. After the experiment you started to wonder whether the coin was fair. 

(a) Formulate test statistic, the null hypothesis and alternative hypothesis for testing the fairness of the coin. 

$$
\begin{aligned}
  p &: \text{Probability of tails} \\
  H_0 &: p = 0.5 \\
  H_1 &: p \neq 0.5 \\
  p_0 &= \frac{7}{11} = `r 7/11`
\end{aligned}
$$

$$
\begin{aligned}
  p-val &= 2\P(t \geq 7/11) \\
  &= `r 2 * pbinom(6, 11, p=0.5, lower.tail=F)`
\end{aligned}
$$

`pbinom(..., lower.tail=F)` is non-inclusive, hence using 6 instead of 7.  
(`#!r 2 * pbinom(6, 11, p = 0.5, lower.tail = FALSE)`)

Could use `binom.test`

```{r}
test <- binom.test(x = 7, n = 11, p = 0.5, alternative = "two.sided")
test$p.value
```

(b) Was the coin fair? Assume 0.05 risk level.

We assume the coin is fair, as the p-value is within the risk level.

## 4. Insurance (R)

In the United States in 1998, the proportion of adults age 21-24 who had no
medical insurance was 34.4 percent, according to the Employee Benefit Research
Institute. A survey of 75 recent college graduates in this age range finds that 40 are
without insurance. Does this support a difference from the nationwide proportion?
Perform a test of significance and report the $p$-value. Is it significant?

```{r}
prop.test(x = 40, n = 70, p = 0.344, alternative = "two.sided", correct = FALSE)
```

The $p$-value is significant.

Could also use a binomial test, since there are only two possible outcomes.  

```{r}
toll_binom <- binom.test(x = 5731, n = 5760, p = 0.999, alternative = "less")
toll_binom
```

This gives a much lower p-value. Because n is low, the normal distribution is
not a great approximation for the binomial distribution.


##  5. Highway Toll (R)

On a number of highways a toll is collected for permission to travel on the
roadway. To lessen the burden on drivers, electronic toll-collection systems are often
used. An engineer wishes to check the validity of one such system. She arranges to
survey a collection unit for single day, finding that of 5,760 transactions, the system
accurately read 5,731. Perform a one-sided significance test to see if this is consistent
with a 99.9% accuracy rating at the 0.05 significance level. (Do you have any doubts that
the normal approximation to the binomial distribution should apply here?)

```{r}
toll_test <- prop.test(x = 5731, n = 5760, p = 0.999,
                       alternative = "less", correct = FALSE)
toll_test
```

This is not consistent with the 99.9% accuracy rating at the 0.05 significance level.
The p-value is $`r toll_test$p.value`$, which is highly significant.

Could also use a binomial test, since there are only two possible outcomes.  
Doesn't change anything though because the p-value is still very low.

```{r}
toll_binom <- binom.test(x = 5731, n = 5760, p = 0.999, alternative = "less")
toll_binom
```

## 6. Car Problems (R)

Historically, a car from a given company has a 10% chance of having a significant
mechanical problem during its warranty period. A new model of the car is being sold. Of
the first 25,000 sold, 2,700 have had an issue. Perform a test of significance to see
whether the proportion of these new cars that will have a problem is more than 10%.
What is the p-value?

```{r}
car_test <- prop.test(x = 2700, n = 25000, p = 0.1,
                      alternative = "greater", correct = FALSE)
car_test
```

The p-value is $`r car_test$p.value`$.

Try binomial test because there are only two possible outcomes.

```{r}
toll_binom <- binom.test(x = 5731, n = 5760, p = 0.999, alternative = "less")
toll_binom
```

Difference is small, because n is large so normal distribution is a good
approximation for the binomial distribution.
